---
title: "Modelling spatial spread I - Exercises"
author: "Andrea Parisi"
output:
  pdf_document: default
  bookdown::html_document2:
    css: style.css
    df_print: paged
  html_document:
    css: style.css
    df_print: paged
  slidy_presentation: default
fontsize: 12pt
---
\sffamily

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r echo=FALSE}
library(pracma)
library(deSolve)
library(viridis)

sir.model <- function( t, x, params )  {
	patches <- params[[1]]   # This is the number of sub-populations
	mu <- params[[2]]        # Birth/death rate
	gamma <- params[[3]]     # Recovery rate
	beta <- params[[4]]      # Common transmission rate
	rho <- params[[5]]       # Contact matrix
	NN  <- params[[6]]       # Sub-population sizes

	# Extract the susceptibles and infectives
	pop.sus <- x[1:patches]
	pop.inf <- x[(patches+1):(2*patches)]
	lambda <- beta * rho %*% pop.inf  # Matrix multiplication

	# Find rhs of (1) using vector calculus
	ds <- mu*(NN-pop.sus)-lambda*pop.sus/NN
	di <- +lambda*pop.sus/NN - (mu+gamma)*pop.inf
	res <- c(ds,di)
	list(res)
}

patches <- 16
NN <- rep(10000, patches)
N0 <- sum(NN)

mu <- 1.0/(70*365)
gamma <- 1./2.2
beta  <- 1.66

randomContactMatrix <- function( patches )  {
	rho <- matrix(2*runif(patches*patches), nrow=patches, ncol=patches) / patches
	# Diagonal elements represent individuals not moving 
	# and are set by applying the constraint on rho
	for (ii in 1:nrow(rho))  {
		rho[ii,ii] <- 0
	}
	# Make sure that sum_j rho_ij = 1
	for (jj in 1:ncol(rho))  {
		if (sum(rho[,jj]) > 1)  {
			rho[,jj] <- rho[,jj] / sum(rho[,jj])
		}
		rho[jj,jj] <- 1-sum(rho[,jj])
	}
	return(rho)
}
rho <- randomContactMatrix(patches)

times <- seq( from = 0, to = 40, by = 0.05)
start.I0 <- rep(0, patches)
start.I0[1] <- 10
start.S0 <- NN-start.I0
xstart <- c( start.S0, start.I0 )
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )

showOutput <- function(out, nregions, sizes)  {
	x.min <- min(out$time)
	x.max <- max(out$time)
	y.min <- 0
	y.max <- max(out[,(2+nregions):ncol(out)])
	plot( c(x.min, x.max), c(y.min, y.max), type="n", col=3, lwd=2,
		  ylim=c(0,y.max), xlab="Time", ylab="Prevalence" )
	for (ii in 1:nregions)  {
		lines( out$time, out[,(1+nregions+ii)], lwd=2, col=viridis(nregions)[ii] )
	}
}
```


## **Exercise 1**

*Increase the number of patches and observe how the infection curves will tend to overlap when such number becomes large.  Can you explain why?*

We start from 16 patches, and we increase the number to 50, 300, 1000.

```{r}
patches <- 16
NN <- rep(10000, patches)

rho <- randomContactMatrix(patches)

start.I0 <- rep(0, patches)
start.I0[1] <- 10
start.S0 <- NN-start.I0
xstart <- c( start.S0, start.I0 )
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```


```{r}
patches <- 50
NN <- rep(10000, patches)

rho <- randomContactMatrix(patches)

start.I0 <- rep(0, patches)
start.I0[1] <- 10
start.S0 <- NN-start.I0
xstart <- c( start.S0, start.I0 )
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```

```{r}
patches <- 300
NN <- rep(10000, patches)

rho <- randomContactMatrix(patches)

start.I0 <- rep(0, patches)
start.I0[1] <- 10
start.S0 <- NN-start.I0
xstart <- c( start.S0, start.I0 )
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```

```{r}
patches <- 1000
NN <- rep(10000, patches)

rho <- randomContactMatrix(patches)

start.I0 <- rep(0, patches)
start.I0[1] <- 10
start.S0 <- NN-start.I0
xstart <- c( start.S0, start.I0 )
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```

EACH patch will average out difference between influx from other patches: as the number of patches increases, all patches will see their influx converge towards a common average value.


# **Exercise 2**

*Increase or decrease the population size and the initial number of infectives of each patch by the same amount.  What happens to the infection curves?  Why?*



```{r}
side <- 4
patches <- 16
NN <- rep(10000, patches)

rho <- randomContactMatrix(patches)

start.I0 <- rep(0, patches)
start.I0[1] <- 10
start.S0 <- NN-start.I0
xstart <- c( start.S0, start.I0 )
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```

```{r}
patches <- 16
NN <- rep(10000000, patches)

rho <- randomContactMatrix(patches)

start.I0 <- rep(0, patches)
start.I0[1] <- 10000
start.S0 <- NN-start.I0
xstart <- c( start.S0, start.I0 )
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```

Modifying the population size does not have any effect since these are continuous model and the contact matrix represents population fractions.


## **Exercise 3**

This is the original nearest-neighbours matrix
```{r}
entry <- 1
rho <- zeros(patches) # Generates a matrix of zeros
for (dst in 1:patches)  {
	for (src in 1:patches)  {
		if (dst == src)  {
			rho[dst,src] <- 0.96 # 1 - 4*0.01 - 4*0.05
		} else if (dst == src+side || dst == src-side ||
				   dst == src+1 || dst == src-1)  {
			rho[dst,src] <- 0.01
		}
	}
}
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```

This is a contact matrix with additional tranmission to second nearest neighbours
```{r}
entry <- 1
rho <- zeros(patches) # Generates a patches x patches matrix of zeros
for (dst in 1:patches)  {
	for (src in 1:patches)  {
		if (dst == src)  {
			rho[dst,src] <- 0.94 # 1 - 4*0.01 - 4*0.05
		} else if (dst == src+side || dst == src-side ||
				   dst == src+1 || dst == src-1)  {
			rho[dst,src] <- 0.01
		} else if (dst == src+side-1 || dst == src-side-1 ||
				   dst == src+side+1 || dst == src-side+1)  {
			rho[dst,src] <- 0.005
		}
	}
}
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```

Increasing the coupling between counties makes outbreaks in far countries occur earlier.


## **Exercise 4**

```{r echo=FALSE}
library(pracma)
library(sf)
#library(lwgeom)  # Required by some shapefiles
library(viridis)

counties.wsg84 <- read_sf(dsn="Data/Local/map.shp")
counties.pop <- read.table(file="Data/Local/population.dat", header=TRUE)
coordinates <- as.matrix( read.table(file="Data/Local/coordinates.dat", 
									 header=TRUE) )
counties.mapper <- read.table(file="Data/Local/mapper.dat", header=TRUE)
patches <- nrow(counties.pop)
comm <- read.table(file="Data/Local/comm.dat", header=TRUE)
```


```{r echo=FALSE}
build.contact.matrix <- function( map, comm, colname)  {
	npatches <- nrow(map)
	mm <- zeros(npatches) # Generates a patches x patches matrix of zeros
	for (dst in 1:npatches)  {
		for (src in 1:npatches)  {
			if (dst == src)  {
				# Entries rho_{jj} will be handled separately
				next
			} else {
				hasentry <- nrow(comm[ comm$dst == dst & comm$src == src, ])
				if (hasentry)  {
					entry <- comm[ comm$dst == dst & comm$src == src, ]
					mm[dst,src] <- entry[,colname]
					mm[dst,src] <- mm[dst,src] / map[ map$id == src, ]$pop
				}
			}
		}
	}
	for (src in 1:npatches)  {
		tot <- sum(mm[,src])
		mm[src,src] <- 1 - tot
	}
	
	## Here a sanity check:
	if (min(mm) < 0)  {
		stop( "Sanity check on rho did not work!")
	}
	return (mm)
}
```

```{r}
NN <- counties.wsg84$pop
beta <- 1.66
times <- seq( from = 0, to = 40, by = 0.05)

rho <- build.contact.matrix( counties.wsg84, comm, "Total" )
patches <- nrow(rho)

# Find counties 'Modoc' and 'Los Angeles'
modoc <- which(counties.wsg84$NAME == "Modoc")
start.I0 <- rep(0, patches)
start.I0[modoc] <- 10
start.S0 <- NN-start.I0
xstart <- c( start.S0, start.I0 )
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```

```{r}
losangeles <- which(counties.wsg84$NAME == "Los Angeles")
start.I0 <- rep(0, patches)
start.I0[losangeles] <- 10
start.S0 <- NN-start.I0
xstart <- c( start.S0, start.I0 )
params <- list(patches, mu, gamma, beta, rho, NN)

res <- deSolve::ode( xstart, times, sir.model, params, method=deSolve::rk4 )

output <- data.frame( res )
showOutput(output, patches, NN)
```

Los Angeles has higher connections than the Modoc county, thus an initial infective in Los Angeles would cause a faster epidemic throughout California.




